Enzo: An Adaptive Mesh Refinement Code for Astrophysics

### 2013-06-07 numerical-radiation-raytracing.tex
Round 2 edits of microphysics & radiation.  There are still missing rates for HD destruction/ionization. Radiation background text is out of sync with the enzo documentation.

%The radiative transfer equation in comoving coordinates
%\citep{Gnedin97} reads
%%
%\begin{equation}
%  \label{eq:rteqn}
%  \frac{1}{c} \; \frac{\partial I_\nu}{\partial t} + 
%  \frac{\hat{n} \cdot \nabla I_\nu}{\bar{a}} -
%  \frac{H}{c} \; \left( \nu \frac{\partial I_\nu}{\partial \nu} -
%  3 I_\nu \right) = -\kappa_\nu I_\nu + j_\nu .
%\end{equation}
%%
%Here $I_\nu \equiv I(\nu, \mathbf{x}, \Omega, t)$ is the radiation
%specific intensity in units of energy per time $t$ per solid angle per
%unit area per frequency $\nu$.  $H = \dot{a}/a$ is the Hubble
%constant, where $a$ is the scale factor.  $\bar{a} = a/a_{em}$ is the
%ratio of scale factors at the current time and time of emission.  The
%second term represents the propagation of radiation, where the factor
%$1/\bar{a}$ accounts for cosmic expansion.  The third term describes
%both the cosmological redshift and dilution of radiation.  On the
%right hand side, the first term considers the absorption coefficient
%$\kappa_\nu \equiv \kappa_\nu(\mathbf{x},\nu,t)$, and the second term
%$j_\nu \equiv j_\nu(\mathbf{x},\nu,t)$ is the emission coefficient
%that includes any point sources of radiation or diffuse radiation.
%GB: this was moved to the equations section.

We solve the radiative transfer equation in comoving coordinates (given by Equation~\ref{eq:rteqn}).
We can make some appropriate approximations to reduce the complexity
of this equation in order to include radiation transport
in numerical calculations.  Typically timesteps in dynamic
calculations are small enough so that $\Delta a/a \ll 1$, therefore
$\bar{a} = 1$ in any given timestep, reducing the second term to
$\hat{n} \partial I_\nu/\partial \mathbf{x}$.  To determine the
importance of the third term, we evaluate the ratio of the third term
to the second term.  This is $HL/c$, where $L$ is the simulation box
length.  If this ratio is $\ll 1$, we can ignore the third term.  For
example at $z=5$, this ratio is 0.1 when $L = c/H(z=5)$ = 53 proper
Mpc.  In large boxes where the light crossing time is comparable to
the Hubble time, then it becomes important to consider cosmological
redshifting and dilution of the radiation.  Thus equation
(\ref{eq:rteqn}) reduces to the non-cosmological form in this local
approximation,
%
\begin{equation}
  \frac{1}{c} \frac{\partial I_\nu}{\partial t} + 
  \hat{n} \frac{\partial I_\nu}{\partial \mathbf{x}} =
  -\kappa_\nu I_\nu + j_\nu .
\end{equation}
%
We choose to represent the source term $j_\nu$ as point sources of
radiation (e.g. stars, quasars) that emit radial rays that are
propagated along the direction $\hat{n}$.

yay tracing is an accurate method to propagate radiation from point
sources on a computational grid, given that there are a sufficient
number of rays passing through each cell.  Along a ray, the radiation
transfer equation reduces to
%
\begin{equation}
\label{eqn:rtray}
\frac{1}{c} \frac{\partial P}{\partial t} + \frac{\partial P}{\partial
  r} = -\kappa P,
\end{equation}

where $P$ is the photon number flux along the ray.  To sample the
radiation field at large radii, ray tracing requires at least $N_{ray}
= 4\pi R^2 / (\Delta x)^2$ rays to sample each cell with one ray,
where $R$ is the radius from the source to the cell and $\Delta x$ is
the cell width.  If one were to trace $N_{ray}$ rays out to $R$, the
cells at a smaller radius $r$ would be sampled by, on average,
$(r/R)^2$ rays, which is computationally wasteful because only a few
rays per cell  are required to provide an accurate calculation
of the radiation field (as we will show later).

We avoid this inefficiency by utilizing adaptive ray tracing
\citep{Abel02_RT}, which progressively splits rays when the sampling
becomes too coarse and is based on Hierarchical Equal Area isoLatitude
Pixelation \citep[HEALPix;][]{HEALPix}.  In this scheme, the rays are
traced along normal directions of the centers of HEALPix pixels, which
evenly divides a sphere into equal areas.  The rays are initialized at
each point source with the photon luminosity (photon s$^{-1}$) equally
spread across $N_{\rm pix} = 12 \times 4^l$ rays, where $l$ is the
initial HEALPix level.  We usually find $l$ = 0 or 1 is sufficient
because these coarse rays will usually be split before traversing
the first cell.

@@

A ray propagates and splits until at least one of the following conditions is met: (1) the photon has traveled $c
\times dt_P$, where $dt_P$ is the radiative transfer timestep; (2) its
photon flux is almost fully absorbed ($>99.9\%$) in a single cell,
which significantly reduces the computational time if the radiation
volume filling fraction is small; (3) the photon leaves the
computational domain with isolated boundary conditions; or (4) the
photon travels $\sqrt{3}$ of the simulation box length with periodic
boundary conditions.

In the first case, the photon is halted at that position and saved,
where it will be considered in the solution of $I_\nu$ at the next
timestep.  In the next timestep, the photon will encounter a different
hydrodynamical and ionization state, hence $\kappa$, in its path.
Furthermore any time variations of the luminosities will be retained
in the radiation field.  This is how this method retains the time
derivative of the radiative transfer equation.  The last restriction
prevents our method from considering sources external to the
computational domain, but a uniform radiation background can be used
in conjunction with ray tracing that adds the local radiation field to
the background intensity.

The radiation field is calculated by integrating Equation
(\ref{eqn:rtray}) along each ray, which is done by considering the
discretization of the ray into segments.  In the following
description, we assume the rays are monochromatic for simplicity.  For
convenience, we express the integration in terms of optical depth
$\tau = \int \kappa(r,t) \; dr$, and for a ray segment,
%
\begin{equation}
  \label{eqn:dtau}
  d\tau = \sigma_{\rm abs}(\nu) n_{\rm abs} dr.
\end{equation}
Here $\sigma_{\rm abs}$ and $n_{\rm abs}$ are the cross section and
number density of the absorbing medium, respectively.  In the static
case, equation (\ref{eqn:rtray}) has a simple exponential analytic
solution, and the photon flux of a ray is reduced by
%
\begin{equation}
  \label{eqn:flux}
  dP = P \times (1 - e^{-\tau})
\end{equation}
as it crosses a cell.  We equate the photo-ionization rate to the
absorption rate, resulting in photon conservation \citep{Abel99_RT,
  Mellema06}.  Thus the photo-ionization and
photo-heating rates associated with a single ray ($k_{\rm ph}$ and
$\Gamma_{\rm ph}$, respectively)  are
%
\begin{equation}
  \label{eqn:kph}
  k_{\rm ph} = \frac{P (1 - e^{-\tau})}{n_{\rm abs} \; V_{\rm cell} \; dt_P},
\end{equation}
\begin{equation}
  \label{eqn:gamma}
  \Gamma_{\rm ph} = k_{\rm ph} \; (E_{\rm ph} - E_i),
\end{equation}
where $V_{\rm cell}$ is the cell volume, $E_{\rm ph}$ is the photon
energy, and $E_i$ is the ionization energy of the absorbing material.
In each cell, the photo-ionization and photo-heating rates from each
ray in the calculation are summed, and after the ray tracing is
complete, these rates are used as inputs to the solver described in
Section~\ref{sec.ov.chem} to update the ionization, chemical, and energy states
of the gas in each cell.

>>>>

%The radiative transfer equation in comoving coordinates
%\citep{Gnedin97} reads
%%
%\begin{equation}
%  \label{eq:rteqn}
%  \frac{1}{c} \; \frac{\partial I_\nu}{\partial t} + 
%  \frac{\hat{n} \cdot \grad I_\nu}{\bar{a}} -
%  \frac{H}{c} \; \left( \nu \frac{\partial I_\nu}{\partial \nu} -
%  3 I_\nu \right) = -\kappa_\nu I_\nu + j_\nu .
%\end{equation}
%%
%Here $I_\nu \equiv I(\nu, \mathbf{x}, \Omega, t)$ is the radiation
%specific intensity in units of energy per time $t$ per solid angle per
%unit area per frequency $\nu$.  $H = \dot{a}/a$ is the Hubble
%constant, where $a$ is the scale factor.  $\bar{a} = a/a_{em}$ is the
%ratio of scale factors at the current time and time of emission.  The
%second term represents the propagation of radiation, where the factor
%$1/\bar{a}$ accounts for cosmic expansion.  The third term describes
%both the cosmological redshift and dilution of radiation.  On the
%right hand side, the first term considers the absorption coefficient
%$\kappa_\nu \equiv \kappa_\nu(\mathbf{x},\nu,t)$, and the second term
%$j_\nu \equiv j_\nu(\mathbf{x},\nu,t)$ is the emission coefficient
%that includes any point sources of radiation or diffuse radiation.
%GB: this was moved to the equations section.

We solve the radiative transfer equation in comoving coordinates (given by Equation~\ref{eq:rteqn}).
We can make some appropriate approximations to reduce the complexity
of this equation in order to include radiation transport
in numerical calculations.  Typically timesteps in dynamic
calculations are small enough so that $\Delta a/a \ll 1$, therefore
$a_{em}/a \approx 1$ in any given timestep, reducing the second term to
$\hat{n} \partial I_\nu/\partial \mathbf{x}$.  To determine the
importance of the third term, we evaluate the ratio of the third term
to the second term.  This is $HL/c$, where $L$ is the simulation box
length.  If this ratio is $\ll 1$, we can ignore the third term.  For
example at $z=5$, this ratio is 0.1 when $L = c/H(z=5)$ = 53 proper
Mpc.  In large boxes where the light crossing time is comparable to
the Hubble time, then it becomes important to consider cosmological
redshifting and dilution of the radiation.  Thus equation
(\ref{eq:rteqn}) reduces to the non-cosmological form in this local
approximation,
%
\begin{equation}
  \frac{1}{c} \frac{\partial I_\nu}{\partial t} + 
  \hat{n} \frac{\partial I_\nu}{\partial \mathbf{x}} =
  -\kappa_\nu I_\nu + j_\nu .
\end{equation}
%
We choose to represent the source term $j_\nu$ as point sources of
radiation (e.g. stars, quasars) that emit radial rays that are
propagated along the direction $\hat{n}$.

Ray tracing is an accurate method to propagate radiation from point
sources on a computational grid as long as there are a sufficient
number of rays passing through each cell.  Along a ray, the radiation
transfer equation reduces to
%
\begin{equation}
\label{eqn:rtray}
\frac{1}{c} \frac{\partial P}{\partial t} + \frac{\partial P}{\partial
  r} = -\kappa P,
\end{equation}
where $P$ is the photon number flux along the ray.  To sample the
radiation field at large radii, ray tracing requires at least $N_{ray}
= 4\pi R^2 / (\Delta x)^2$ rays to sample each cell with one ray,
where $R$ is the radius from the source to the cell and $\Delta x$ is
the cell width.  If one were to trace $N_{ray}$ rays out to $R$, the
cells at a smaller radius $r$ would be sampled by, on average,
$(r/R)^2$ rays, which is computationally wasteful because only a few
rays per cell  are required to provide an accurate calculation
of the radiation field (as we will show later).

We avoid this inefficiency by utilizing adaptive ray tracing
\citep{Abel02_RT}, which is based on Hierarchical Equal Area isoLatitude
Pixelation \citep[HEALPix;][]{HEALPix} and progressively splits rays when the sampling
becomes too coarse.  In this approach, the rays are
traced along normal directions of the centers of the HEALPix pixels that 
evenly divide a sphere into equal areas.  The rays are initialized at
each point source with the photon luminosity (photon s$^{-1}$) equally
spread across $N_{\rm pix} = 12 \times 4^l$ rays, where $l$ is the
initial HEALPix level.  We usually find $l$ = 0 or 1 is sufficient
because these coarse rays will usually be split before traversing
the first cell.

@@

A ray propagates and splits until at least one of the following conditions is met: (1) the photon has traveled $c
\times dt_P$, where $dt_P$ is the radiative transfer timestep; (2) its
photon flux is almost fully absorbed ($>99.9\%$) in a single cell,
which significantly reduces the computational time if the radiation
volume filling fraction is small; (3) the photon leaves the
computational domain with isolated boundary conditions; or (4) the
photon travels $\sqrt{3}$ of the simulation box length with periodic
boundary conditions.
In the first case, the photon is halted at that position and saved,
where it will be considered in the solution of $I_\nu$ at the next
timestep.  In the next timestep, the photon will encounter a different
hydrodynamical and ionization state, hence $\kappa$, in its path.
Furthermore any time variations of the luminosities will be retained
in the radiation field.  This is how this method retains the time
derivative of the radiative transfer equation.  The last restriction
prevents our method from considering sources external to the
computational domain. However, a uniform radiation background can be used
in conjunction with ray tracing that adds the background intensity to the
local radiation field.

The radiation field is calculated by integrating Equation
(\ref{eqn:rtray}) along each ray, which is done by considering the
discretization of the ray into segments.  In the following
description, we assume the rays are monochromatic for simplicity.  For
convenience, we express the integration in terms of optical depth
$\tau = \int \kappa(r,t) \; dr$, and for a ray segment,
%
\begin{equation}
  \label{eqn:dtau}
  d\tau = \sigma_{\rm abs}(\nu) n_{\rm abs} dr.
\end{equation}
Here $\sigma_{\rm abs}$ and $n_{\rm abs}$ are the cross section and
number density of the absorbing medium, respectively.  In the static
case, equation (\ref{eqn:rtray}) has a simple exponential analytic
solution, and the photon flux of a ray is reduced by
%
\begin{equation}
  \label{eqn:flux}
  dP = P \times (1 - e^{-\tau})
\end{equation}
as it crosses a cell.  We equate the photo-ionization rate to the
absorption rate, resulting in photon conservation \citep{Abel99_RT,
  Mellema06}.  Thus the photo-ionization and
photo-heating rates associated with a single ray ($k_{\rm ph}$ and
$\Gamma_{\rm ph}$, respectively)  are
%
\begin{equation}
  \label{eqn:kph}
  k_{\rm ph} = \frac{P (1 - e^{-\tau})}{n_{\rm abs} \; V_{\rm cell} \; dt_P},
\end{equation}
\begin{equation}
  \label{eqn:gamma}
  \Gamma_{\rm ph} = k_{\rm ph} \; (E_{\rm ph} - E_i),
\end{equation}
where $V_{\rm cell}$ is the cell volume, $E_{\rm ph}$ is the photon
energy, and $E_i$ is the ionization energy of the absorbing material.
In each cell, the photo-ionization and photo-heating rates from each
ray in the calculation are summed. After the ray tracing is
complete, these rates are used as inputs to the solver described in
Section~\ref{sec.ov.chem} to update the ionization, chemical, and energy states
of the gas in each cell.
